New approaches for boosting to uniformity

### 2014-09-17 assets/paperdraft/paper_draft.tex

\author{
Aleksandar Bukva$^{\rm ADD}$, 
Vladimir Gligorov$^d$,
Alex Rogozhnikov$^{a,b}$\thanks{Corresponding author.}~ ,
Andrey Ustuzhanin$^b$ and
Mike Williams$^c$\\
\llap{$^a$}Lomonosov Moscow State University, Moscow, Russia\\
\llap{$^b$}Yandex LLC, Moscow, Russia\\
\llap{$^c$}Massachusetts Institute of Technology, Cambridge, MA, United States \\
\llap{$^d$}Organisation Europ\'eenne pour la Recherche Nucl\'eaire (CERN), Geneva, Switzerland  \\
E-mail: \email{alex.rogozhnikov@yandex.ru}}

@@

% AMS packages:
\usepackage{amsmath, amsthm, amsfonts, xspace}

@@

\def \x {\ensuremath{\vec{x}}\xspace}
\def \y {\ensuremath{\vec{y}}\xspace}
\def\knn{\text{knn}}
\def\knni{\text{knn}(i)}

@@

%\section{Uniformity Measurement}
%\input{parts/metrics}
%\section{Approaches Proposed}
%\input{parts/approaches}

\section{Uniformity Boosting Methods}
\input{parts/methods}

@@

ADD THEIL, ETC DESCRIPTIONS AND HOW TO IMPLEMENT IN THE ALGORITHM HERE.

>>>>

\author{
Aleksandar Bukva$^e$, 
Vladimir Gligorov$^d$,
Alex Rogozhnikov$^{a,b}$\thanks{Corresponding author.}~ ,
Andrey Ustuzhanin$^b$ and
Mike Williams$^c$\\
\llap{$^a$}Lomonosov Moscow State University, Moscow, Russia\\
\llap{$^b$}Yandex LLC, Moscow, Russia\\
\llap{$^c$}Massachusetts Institute of Technology, Cambridge, MA, United States \\
\llap{$^d$}Organisation Europ\'eenne pour la Recherche Nucl\'eaire (CERN), Geneva, Switzerland  \\
\llap{$^e$}Faculty of Physics, Belgrade, Serbia \\
E-mail: \email{alex.rogozhnikov@yandex.ru}}

@@

% AMS packages:

\usepackage{amsmath, amsthm, amsfonts, xspace}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{float}

@@

\def\x {\ensuremath{\vec{x}}\xspace}
\def\y {\ensuremath{\vec{y}}\xspace}
\def\knn{\text{knn}}
\def\knni{\text{knn}(i)}

@@

\section{Uniformity Boosting Methods}
\input{ugb}

@@

Clean all this up...

%\section{Uniformity Measurement}
\input{parts/metrics}
%\section{Approaches Proposed}
\input{parts/approaches}

### 2014-09-17 assets/paperdraft/parts/metrics.tex

\subsection{Advantages and Disadvantages of Different Metrics}

... TODO, here some plots with comparison, maybe timings ...

>>>>

\subsection{Advantages and Disadvantages of Different Metrics}
Let us compare two metrics that have shown most appropriate for our problem, SDE and Theil. We have some linearly spaced array of mass of particles
from interval $[0,1]$, some constant $\alpha$ linearly sapced in interval $[.1,1]$. The predictions are correlated with mass via beta distribution.
We have two distributions that are different in a way that they are symmetrical. SDE should show no changes if we flip the distribution, while the Theil should make difference between pit and peak. First distribution with a peak is given by:

\[
	p\text{[i]}=\textbf{beta}(\alpha \cdot \bar m, \text{abs}(m\text[i]-\bar m))
\]

while the second one has minus sign in front.
\begin{figure}[H]
\centering
		\begin{subfigure}[b]{0.95\textwidth}
			\includegraphics[width=\textwidth]{graphs/distribution1.png}
			\caption{Mass vs prediction}
		\end{subfigure}
		\begin{subfigure}[b]{0.95\textwidth}
			\includegraphics[width=\textwidth]{graphs/efficiencies1.png}
			\caption{Efficiencies}
		\end{subfigure}
		\caption{Peak distribution}
\end{figure}

\begin{figure}[H]
\centering
	\includegraphics[width=8cm]{graphs/measures1.png}
	\caption{SDE and Theil}
\end{figure}

\begin{figure}[H]
\centering
		\begin{subfigure}[b]{0.95\textwidth}
			\includegraphics[width=\textwidth]{graphs/distribution2.png}
			\caption{Mass vs prediction}
		\end{subfigure}
		\begin{subfigure}[b]{0.95\textwidth}
			\includegraphics[width=\textwidth]{graphs/efficiencies2.png}
			\caption{Efficiencies}
		\end{subfigure}
		\caption{Pit distribution}
\end{figure}

\begin{figure}[H]
\centering
	\includegraphics[width=8cm]{graphs/measures2.png}
	\caption{SDE and Theil}
\end{figure}

From these figures we can see that SDE has the same value and it is not able to make difference between these distributions. In the second case Theil has lower value which indicates that distribution is flatter. In the combination with SDE it can be used to detarmine the type of a distribution. Lower value of SDE can indicate that distribution is flat or has some narrow peaks or pits, and by using Theil we can distinguish between these two cases.

### 2014-09-17 assets/paperdraft/ugb.tex

Traditional boosting works by assigning training events more weight based on classification erros made by previous members of the series.  For example, in gradient boosting~\cite{ref:gradboost} the analyst chooses a differentiable loss function with the goal of building a classifier that minimizes the loss.  
A popular choice of loss function is the AdaBoost~\cite{ref:adaboost} loss function defined as follows:
\begin{equation}
L_{\rm ada} = \sum\limits_{i=1}^{n}w_i \times {\rm exp}\left[-\gamma_i s_i\right], 
\end{equation}
where $w$ is the weight of each event\footnote{We use the convention that $\sum w_i = 1$ for both signal and background events at each stage of training.}, $\gamma = +1(-1)$ for signal(background) events and $s$ is the so-called score of each event which is obtained as the sum of scores of all of the previous classifiers in the series.   
At each stage in the gradient boosting process, a {\em regressor} (a decision tree in our case) is trained whose purpose is to decrease the loss.  This is accomplished using the gradient decent method and the {\em pseudo-residuals}
\begin{equation}
  -\frac{\partial L_{\rm ada}}{\partial s_i} = w_i \gamma_i  {\rm exp}\left[-\gamma_i s_i\right],
\end{equation}
which are positive(negative) for signal(background) events and have larger modulai for poorly classified events.

The gradient-boosting algorithm is general in that it only requires the analyst specify a loss function and its gradient.  The AdaBoost loss function considers each event individually, but can easily be modified to take into account {\em non-local} properties of the classifier as follows:
\begin{equation}
  \label{eq:loss_general}
  L_{\rm general} = \sum\limits_{i=1}^{n} w_i {\rm exp}\left[-\gamma_i \sum\limits_{j} a_{ij} s_j \right], 
\end{equation} 
where $A$ is a matrix\footnote{A natural choice is a square $n\times n$ matrix, but this is not required.}.  For the case where $A$ is the identity matrix, the AdaBoost loss function is recovered.  
Other choices of $A$ will induce non-local effects, {\em e.g.}, consider the sparse matrix 
\begin{equation}
  a_{ij}^{\rm knn} = \begin{cases} 
    1, & j \in \knni, \text{ events $i$ and $j$ belong to the same class} \\
    0, & \text{otherwise},
\end{cases}
\end{equation}
where $\knn(i)$ denotes the set of $k$-nearest-neighbor events to event $i$.   The loss function obtained from Eq.~\ref{eq:loss_general} using $A^{\rm knn}$, denoted by $L_{\rm knn}$, accounts for the score of each event's $k$ nearest neighbors and not just each event individually.
The pseudo-residuals are then
\begin{equation}
  -\frac{\partial L_{\rm knn}}{\partial s_k} = -\sum\limits_i w_i \gamma_i a_{ik} {\rm exp}\left[-\gamma_i \sum_j a_{ij} s_j \right].
\end{equation}
One can see that the direction of the gradient will be influenced the most by events whose $k$-nearest-neighbor events are classified poorly.

Another approach is to include in the definition of the loss function some uniformity metric.  Consider first the case where the data has been binned in \y.  If the distribution of classifier responses in each bin, $f_b(s)$, is the same as the global response distribution, $f(s)$, then any cut made on the response will produce a uniform selection efficiency in \y.  Therefore, performing a one-dimensional goodness-of-fit test of the hypothesis $f_b \equiv f$ in each bin provides an assessment of the selection uniformity.  
For example, one could perform the Kolmogorov-Smirnov test in each bin and define a loss function as follows:
\begin{equation}
  L_{\rm flat(KS)} = \sum\limits_{b} w_b {\rm max}|F_b(s)-F(s)|,
\end{equation}
where $F_{(b)}(s)$ denotes the cummulative distribution of $f_{(b)}(s)$ and $w_b = \sum \delta({\rm bin}_i - b) w_i$, {\em i.e.} $w_b$ is the sum of the weights of the signal events in the bin.

The gradient of the Kolmogorov-Smirnov loss function is zero for events with responses greater than the value of $s$ at which ${\rm max}|F_b(s)-F(s)|$ occurs.  Thus, it is not suitable for gradient boosting due to its instability.  Instead, we use the following loss function:
\begin{equation}
  L_{\rm flat} = \sum\limits_{b} w_b \sum\limits_{i \in b}|F_b(s_i)-F(s_i)|,
\end{equation}
whose pseudo-residuals are
\begin{equation}
 -\frac{\partial L_{\rm flat}}{\partial s_k} = - w_b {\rm sign}\left[F_b(s_k)-F(s_k)\right] \left( w_k/w_b - w_k \right).  
\end{equation}
This so-called flatness loss penalizes non-uniformity but does not consider the quality of the classification.  Therefore, the full loss function used is 
\begin{equation}
\label{eq:L_ada_flat}
  L_{{\rm ada}+{\rm flat}} = L_{\rm flat} + \alpha L_{\rm ada},
\end{equation}  
where $\alpha$ is a real-valued parameter that is typically chosen to be small.  The first term in Eq.~\ref{eq:L_ada_flat} penalizes non-uniformity, while that second term penalizes poor classification.

ADD KNN HERE ONCE ALGORITHM IS TESTED.

>>>>

One category of boosting works by assigning training events more weight based on classification erros made by previous members of the series.  For example, the AdaBoost~\cite{ref:adaboost} algorithm updates the weight of event $i$, $w_i$, according to
\begin{equation}
w_i^{\prime} = w_i \times {\rm exp}\left[-\gamma_i s_i\right], 
\end{equation}
where $\gamma = +1(-1)$ for signal(background) events and $s$ is the so-called score of each event which is obtained as the sum of scores of all of the previous classifiers in the series.  
The uBoost technique, described in detail in Ref.~\cite{cit:uboost}, alters the event-weight updating procedure to achieve uniformity in the signal-selection efficiency.

Another approach to obtain uniformity, introduced in this paper, involves defining a more general expression of the AdaBoost criteria:
\begin{equation}
 w_i^{\prime} = w_i \times {\rm exp}\left[-\gamma_i \sum_j a_{ij}\right],
\end{equation}
where $a_{ij}$ are the elements of some matrix $A$\footnote{A natural choice is a square $n\times n$ matrix, but this is not required.}.  For the case where $A$ is the identity matrix, the AdaBoost weighting procedure is recovered.  
Other choices of $A$ will induce non-local effects, {\em e.g.}, consider the sparse matrix $A_{\rm knn}$ given by
\begin{equation}
  a_{ij}^{\rm knn} = \begin{cases} 
    \frac{1}{k}, & j \in \knni, \text{ events $i$ and $j$ belong to the same class} \\
    0, & \text{otherwise},
\end{cases}
\end{equation}
where $\knn(i)$ denotes the set of $k$-nearest-neighbor events to event $i$.
This procedure for updating the event weights, which we refer to as kNNAdaBoost, accounts for the score of each event's $k$ nearest neighbors and not just each event individually.

The gradient boosting~\cite{ref:gradboost} (GB) algorithm category requires the analyst to choose a differentiable loss function with the goal of building a classifier that minimizes the loss.  
A popular choice of loss function is the AdaLoss~\cite{ref:adaboost} function 
\begin{equation}
L_{\rm ada} = \sum\limits_{i=1}^{n} {\rm exp}\left[-\gamma_i s_i\right]. 
\end{equation}
At each stage in the gradient boosting process, a {\em regressor} (a decision tree in our case) is trained whose purpose is to decrease the loss.  This is accomplished using the gradient decent method and the {\em pseudo-residuals}
\begin{equation}
  -\frac{\partial L_{\rm ada}}{\partial s_i} = \gamma_i  {\rm exp}\left[-\gamma_i s_i\right],
\end{equation}
which are positive(negative) for signal(background) events and have larger modulai for poorly classified events.

The gradient-boosting algorithm is general in that it only requires the analyst specify a loss function and its gradient.  The AdaLoss function considers each event individually, but can easily be modified to take into account non-local properties of the classifier as follows:
\begin{equation}
  \label{eq:loss_general}
  L_{\rm general} = \sum\limits_{i=1}^{n} {\rm exp}\left[-\gamma_i \sum\limits_{j} a_{ij} s_j \right]. 
\end{equation} 
For example, the loss function obtained from Eq.~\ref{eq:loss_general} using $A_{\rm knn}$, which we refer to as kNNAdaLoss and denote $L_{\rm knn}$, accounts for the score of each event's $k$ nearest neighbors and not just each event individually.
The pseudo-residuals of $L_{\rm knn}$ are 
\begin{equation}
  -\frac{\partial L_{\rm knn}}{\partial s_k} = -\sum\limits_i \gamma_i a_{ik}^{\rm knn} {\rm exp}\left[-\gamma_i \sum_j a_{ij}^{\rm knn} s_j \right].
\end{equation}
One can see that the direction of the gradient will be influenced the most by events whose $k$-nearest-neighbor events are classified poorly.

Another approach is to include in the definition of the loss function some uniformity metric.  Consider first the case where the data have been binned in \y.  If the distribution of classifier responses in each bin, $f_b(s)$, is the same as the global response distribution, $f(s)$, then any cut made on the response will produce a uniform selection efficiency in \y.  Therefore, performing a one-dimensional goodness-of-fit test of the hypothesis $f_b \equiv f$ in each bin provides an assessment of the selection uniformity.  
For example, one could perform the Kolmogorov-Smirnov test in each bin and define a loss function as follows:
\begin{equation}
  L_{\rm flat(KS)} = \sum\limits_{b} w_b {\rm max}|F_b(s)-F(s)|,
\end{equation}
where $F_{(b)}(s)$ denotes the cummulative distribution of $f_{(b)}(s)$ and $w_b = \sum \delta({\rm bin}_i - b)/n_{\rm signal}$, {\em i.e.} $w_b$ is the fraction of signal events in the bin\footnote{If weighted events are used, then the fractional sum of weights should be used for $w_b$.}.

The gradient of the Kolmogorov-Smirnov loss function is zero for events with responses greater than the value of $s$ at which ${\rm max}|F_b(s)-F(s)|$ occurs.  Thus, it is not suitable for gradient boosting due to its instability.  Instead, we use the following {\em flatness} loss function:
\begin{equation}
  L_{\rm flat} = \sum\limits_{b} w_b \int |F_b(s)-F(s)|^2 ds,
\end{equation}
whose pseudo-residuals are
\begin{equation}
 -\frac{\partial L_{\rm flat}}{\partial s_k} \approx - 2\left[F_b(s_k)-F(s_k)\right]/n_{\rm signal}.
\end{equation}
This so-called flatness loss penalizes non-uniformity but does not consider the quality of the classification.  Therefore, the full loss function used is 
\begin{equation}
\label{eq:L_ada_flat}
  L_{{\rm ada}+{\rm flat}} = L_{\rm flat} + \alpha L_{\rm ada},
\end{equation}  
where $\alpha$ is a real-valued parameter that is typically chosen to be small.  The first term in Eq.~\ref{eq:L_ada_flat} penalizes non-uniformity, while that second term penalizes poor classification.

The loss function given in Eq.~\ref{eq:L_ada_flat} can also be constructed without binning the data using k-nearest-neighbor events.  The cummulative distribution $F_{\rm knn}(s)$ is easily obtained and the bin weight, $w_b$, is replaced by ???
